---
title: "Airquality_cleaning"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Data Cleaning with R

## Content Summary

1. Exploring dataframe
2. Data types
3. Plotting to view correlations and for outlier identification
4. Identifying and dealing with NAs

```{r airquality}
# Let us use a dataset that is availabe within base R

data_to_clean <- airquality

# We will start by exploring this data set in order to know various attributes about it. For instance
# how many rows and columns does it have, what are the names of the columns and what are their data types.

# Number of columns and rows
length(data_to_clean)
ncol(data_to_clean)

length(data_to_clean[,1])
nrow(data_to_clean)

dim(data_to_clean)

# Names of columns
names(data_to_clean)
```


```{r}

# Structure of dataframe
str(data_to_clean)

dplyr::glimpse(data_to_clean)

```

In the case where you want to know the data type of the object you are working on

```{r}
 
class(data_to_clean)

class(data_to_clean$Wind)
class(data_to_clean$Month)

```
```{r}
# Note that month has been categorised as type integer. We would like this to be factor or categorical.

data_to_clean$Month <- as.factor(data_to_clean$Month)
class(data_to_clean$Month)

```
Sometimes it is necessary to leave the categorical variables with numerical values. However, let's say that in
our case, we would like the variable month to be indicated by the month name. Let's change this.
```{r}
levels(data_to_clean$Month) # To see the levels. Notice that they are not stored as integers but as string

levels(data_to_clean$Month) <- list(May = "5", June="6", July="7", August="8", September = "9")
levels(data_to_clean$Month)
```


```{r}
# We may want to know the number of observation per category of a variable. In our instance use could use Month.
xtabs(~data_to_clean$Month, data = data_to_clean)

# To present this data into a data frame simply...
as.data.frame(xtabs(~data_to_clean$Month, data = data_to_clean))
```

## Using plotting to explore data and detect outliers
Part of exploring data is identifying relationships between the variables in our dataframe. Scatter plots help us to visualize data and detect possible correlations.

```{r}
plot(data_to_clean)
```
Are there any variables that seem to be correlated?

Let us focus on the relationship between Ozone and Temperature.
```{r}

attach(data_to_clean) # Fun tip: When using one data set it can be convenient to store the variables in working memory so that you don't have to use '$' all through. Remember to detach() to remove it from working memory.

plot(Ozone, Temp)
```
We can see outliers here. It is possible to determine those outliers from the scatter plot.

```{r}
attach(data_to_clean)
plot(Ozone, Temp)
outlier_index <- identify(data_to_clean$Ozone, data_to_clean$Temp) # This allows you to click on points of interest on a plot and store their index.

# Click the outliers and once done click FINISH.
# The points will be indicated by their index values.
# To label the points with other values, for instance, the month
plot(Ozone, Temp)
outlier_index <- identify(Ozone, Temp, labels = Month, plot=TRUE)

# Dealing with outliers is beyond the scope of this lab session. :-)

# paste(as.character(Date), "_", as.character(Month))
detach(data_to_clean)

```
 Now let's try some summary stats
```{r}

summary(data_to_clean)


```
 Here we get the summary stats. However, notice that there are NA values. How do NA values affect analysis?
 
```{r}
1+NA

mean(data_to_clean$Ozone)
```
So how does summary() find the mean then?
```{r}
mean(data_to_clean$Ozone, na.rm = TRUE)
```

